---
title: "Mortgage Lending Analysis_EDA"
author: "Florence Yan"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: '3'
  html_notebook:
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
---

```{r echo=FALSE}
# Clear environment of variables and functions
rm(list = ls(all = TRUE)) 
# Clear environmet of packages
if(is.null(sessionInfo()$otherPkgs) == FALSE)lapply(paste("package:", names(sessionInfo()$otherPkgs), sep=""), detach, character.only = TRUE, unload = TRUE)
```

```{r message = FALSE, echo = FALSE, error = FALSE}
library(stargazer)
library(tidyverse)
library(ggplot2)
library(GGally)
library(gridExtra)
library(kableExtra)
library(stargazer)
library(stargazer)
library(lmtest) # for BPG test
library(sandwich) # for robust standard error
library(aod)
library(Rcpp)
```


```{r}
# Load data
MLD_data <- read.csv('MLD DATA File.csv')
```


```{r}
# Mutate loan/purchase into percantage
MLD_data <- MLD_data %>% 
  mutate(LOANPRC_percent = LOANPRC*100)

#Select columns
MLD_data <- MLD_data[c('MARRIED','GDLIN','OBRAT','BLACK','HISPAN','MALE','APPROVE','LOANPRC_percent')]
```

# Research Question 

  + Controlling for relevant characteristics, is race/ethnicity associated with the outcome of a mortgage loan application? 


# Data Description 

  + APPROVE = 1 if mortgage loan was approved, = 0 otherwise

  + GDLIN = 1 if credit history meets guidelines, = 0 otherwise

  + LOANPRC = loan amount/purchase price

  + OBRAT = other obligations as a percent of total income

  + MALE = 1 if male, = 0 otherwise

  + MARRIED = 1 if married, = 0 otherwise

  + BLACK = 1 if black, = 0 otherwise

  + HISPAN = 1 if Hispanic, = 0 otherwise
  
  + All applicants are non-Hispanic white, non Hispanic black, or Hispanic

# Data Cleaning
## Change Data Types
## Change Column Names

```{r}
# Change the type of variables
MLD_data$GDLIN <- as.factor(MLD_data$GDLIN)
MLD_data$BLACK <- as.factor(MLD_data$BLACK)
MLD_data$HISPAN <- as.factor(MLD_data$HISPAN)
MLD_data$APPROVE <- as.factor(MLD_data$APPROVE)

colnames(MLD_data) <- c("Marrige_Status", "Credit_Meet", "Obliga_TotIncome", "Black", "Hispanic", "Gender", "Approve", "Loan_Purchase")

str(MLD_data)
kable(summary(MLD_data))
```

# Findings from summary
+ There are 6 factors, 2 are numerical
+ There are missing values in Marrige Status and Gender
+ There are non-sense values in Credit History('666')
+ Most of our samples are non-black and non-hispanic males
+ 

# Filtering
```{r}
# Eliminate N/A
MLD <-
  MLD_data %>%
  filter(Credit_Meet != 666) %>%
  filter(Gender != ".") %>%
  filter(Marrige_Status != ".") 
  
attach(MLD)
str(MLD)
kable(summary(MLD))
```
```{r}
# Present descriptive statistics for the sample overall and separately for 
# non-Hispanic white, non-Hispanic black, and Hispanic loan applicants.

## overall descriptive analysis
stargazer(MLD, title = "Table 1: Data Distribution", type = "text")
head(MLD)

## descriptive analysis in different races
MLD %>% 
  select(Obliga_TotIncome, Black, Hispanic)%>%
  mutate(race = ifelse(Black == 0 & Hispanic == 0, "White", ifelse(Black == 1 & Hispanic == 0, "Black", "Hispanic" ))) %>%
  group_by(race) %>%
  mutate(id = 1:n()) %>% 
  as.data.frame() %>% 
  spread(race, Obliga_TotIncome) %>% 
  select(-id) %>% 
  stargazer(digits = 0, header = F, 
            title = "Table 3: Loan_purchase data distribution", 
            type = "text")

MLD %>% 
  select(Loan_Purchase, Black, Hispanic)%>%
  mutate(race = ifelse(Black == 0 & Hispanic == 0, "White", ifelse(Black == 1 & Hispanic == 0, "Black", "Hispanic" ))) %>%
  group_by(race) %>%
  mutate(id = 1:n()) %>% 
  as.data.frame() %>% 
  spread(race, Loan_Purchase) %>% 
  select(-id) %>% 
  stargazer(digits = 0, header = F, 
            title = "Table 3: Loan_purchase data distribution", 
            type = "text")

```

# Correlations
```{r error=FALSE, warning=FALSE, message=FALSE}
ggpairs(MLD,lower=list(combo=wrap("facethist",  
binwidth=40)))
```

# Histograms of quantitive variables
```{r}
grid.arrange(
  MLD_data %>% 
    ggplot(aes(Obliga_TotIncome)) + 
    geom_histogram(binwidth=8, fill="black"),
  MLD_data %>% 
    ggplot(aes(Loan_Purchase)) +
    geom_histogram(binwidth=8, fill="black"),
  ncol = 2
)
```

# Use tile plot to show multivariate categorical

```{r}
MLD %>%
  group_by(Credit_Meet, Approve) %>% 
  summarise(count = n()) %>% 
  ggplot(aes(Credit_Meet, Approve)) + 
  geom_tile(aes(fill = -count))
```

+ Most cases are approved when applicates' credit histories meet guiedlines


# Differenciate ethnicities
```{r}
# Black
MLDB <-
  MLD %>%
  filter(Black == 1) 
# Hispanic
MLDH <-
  MLD %>%
  filter(Hispanic == 1)
# White
MLDW <-
  MLD %>%
  filter(Black == 0) %>%
  filter(Hispanic == 0)
```

# Summary for each ethnicity
```{r}
kable(summary(MLDB))
kable(summary(MLDH))
kable(summary(MLDW))
```


## Race/ethnicity might associated with the outcome of a mortgage loan application

  - White group has the highest mean of Approve
  - White group has the highest mean of Credit_History_Meets_Guidlines


# Logit Model
```{r}
#Estimate Logit Model
LogitModel = glm(Approve ~ Marrige_Status + Credit_Meet + Obliga_TotIncome + Black + Hispanic + Gender + Loan_Purchase, data = MLD, 
                 family = "binomial")

summary(LogitModel)

#Generate Odds Ratios
exp(coef(LogitModel))


# #Define prototypical loan applicants (you will need more than 3)
# prototype1 <- data.frame(OBRAT=mean(MLD$OBRAT),BLACK = 1, HISPAN = 0)
# prototype2 <- data.frame(OBRAT=mean(MLD$OBRAT),BLACK = 0, HISPAN = 1)
# prototype3 <- data.frame(OBRAT=mean(MLD$OBRAT),BLACK = 0, HISPAN = 0)
# 
# #Predict probabilities for prototypical individuals
# prototype1$predictedprob <- predict (LogitModel, newdata = prototype1, type ="response")
# prototype2$predictedprob <- predict (LogitModel, newdata = prototype2, type ="response")
# prototype3$predictedprob <- predict (LogitModel, newdata = prototype3, type ="response")
# 
# prototype1
# prototype2
# prototype3
```


# Probit Model

```{r}
#Estimate Probit Model
ProbitModel = glm(Approve ~ Marrige_Status + Credit_Meet + Obliga_TotIncome + Black + Hispanic + Gender + Loan_Purchase, data = MLD, 
                  family = "binomial" (link = "probit"))
summary(ProbitModel)

# #Predict probabilities for prototypical individuals
# prototype1$predictedprob <- predict (ProbitModel, newdata = prototype1, type ="response")
# prototype2$predictedprob <- predict (ProbitModel, newdata = prototype2, type ="response")
# prototype3$predictedprob <- predict (ProbitModel, newdata = prototype3, type ="response")
# 
# prototype1
# prototype2
# prototype3
```

